Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-13 _tex/paper2.tex
Built site for gh-pages

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The new model integrates the
theoretical core principles alongside key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. Section~\ref{sec-statistical} translates these
theoretical and practical elements into a probabilistic statistical
model for the analysis of pairwise comparison data.
Section~\ref{sec-discussion} discusses the implications of the findings
and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's key insights.

@@

\section{Extending Thurstone's general form}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed development. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this,
researchers could design a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges would then compare pairs of students' written texts (stimuli), to
produce a dichotomous outcome that reflects the relative quality of each
text (trait). Based on this setup, researchers could reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Finally, following current CJ practices,
researchers would then rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify various
estimands relevant to CJ inquiries, including the one described in the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. \emph{Consistency} refers
to the property of an estimator whose estimates converge to the ``true''
value of the estimand as the data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency cannot
be achieved, even with ``infinite'' and error-free data. Consequently,
deriving meaningful insights from finite data becomes impossible
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

Thus, this section addresses the identified issues in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
leverages the capabilities of this approach to formalize the combination
of the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to a conceptual
population of comparison data, that is, data representing all repeated
comparisons made by every available judge for each pair of stimuli
produced by each pair of individuals in the population, hence its name.
Conversely, Section~\ref{sec-theory-theoretical_SC} integrates the
assessment design features into what we call the \emph{sample-comparison
model}. This model assumes a more realistic scenario where researchers
only have access to a sample of judges, individuals, stimuli, and
comparisons from the conceptual population.

@@

Figure~\ref{fig-cj04} illustrates the preliminary integration of
theoretical principles into the conceptual-population model.
Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
motivating example. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
variables. In this SCM, the subscripts \(\{a,b\}\) represent the
compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
label the students who wrote the texts (individuals), where \(h = i\) is
possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
the judgment number, which accounts for experimental cases where judges
perform multiple judgments of the same texts written by the same
students, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}.

The other variables in SCM~\ref{fig-cj04_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
presented in Section~\ref{sec-theory-issue1b} and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
judges' biases \((B_{kj})\) into the model. Together with the outcome,
these variables form the preliminary set of endogenous variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

By incorporating judges' biases into the conceptual-population model, we
assume that additional relevant characteristics of the stimuli are
unlikely to be equally weighted or opposing. Thus, these characteristics
may exert an uneven influence on judges' perceptions, creating biases in
their judgments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases may ultimately introduce dependencies between
the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}, hence the need
for their integration. The full statistical elaboration of this point is
presented in Section~\ref{sec-statistical}.

@@

\[
\begin{aligned}
  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
\end{aligned}
\]

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering the advantage of
highlighting potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}. Thus, this study uses the magnified
representation.

Figure~\ref{fig-cj04_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ experiments. It also distinguishes between observable
endogenous variables, such as the outcome, and variables considered
latent according to the theory and practice of CJ, such as the the
text's discriminal processes, their discriminal difference, and the
judges' biases.

It is easy to see that representing the CJ system using indices for a
single pair of individuals and stimuli, along with the indices for the
judgment number and the judge responsible for the comparison, can
quickly become intractable and cumbersome. Thus, to maintain generality
and simplify the notation, certain indices are grouped into vectors or
matrices. First, we define the matrices \(IA\) and \(JK\) as:

\[
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
\]

Each row of \(IA\) represents a unique pairing of an individual
\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
number of individuals and \(n_{A}\) the number of stimulus within
individuals in the population. Similarly, each row of \(JK\) represents
a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
represents the number of judges and \(n_{K}\) the number of judgments
performed by each judge in the population.

@@

@@

>>>>

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} develops Thurstone's theory to
address these challenges. The development integrates core theoretical
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the implications of the findings and explores avenues for
future research. Finally, Section~\ref{sec-conclusion} summarizes the
study's key insights.

@@

\section{Developing Thurstone's theory}\label{sec-theoretical}

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed development. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this,
researcher design a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli), to
produce a dichotomous outcome that reflects the relative quality of each
text (trait). Based on this setup, researchers reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Finally, following current CJ practices, researchers
then rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that Thurstone's Case V, and by extension the BTL model, suffers from
several statistical and measurement limitations. These limitations
hinder the model's ability to identify various estimands relevant to CJ
inquiries, including the one described in the motivating example.
Identification is crucial because it is a necessary condition for
ensuring consistent estimators. \emph{Consistency} refers to the
property of an estimator whose estimates converge to the ``true'' value
of the estimand as the data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency cannot
be achieved, even with ``infinite'' and error-free data. Thus, deriving
meaningful insights from finite data becomes impossible
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by developing Thurstone's theory using
the structural approach to causal inference. Specifically, it leverages
the capabilities of this approach to formalize the combination of the
core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to comparison
data from a conceptual population, that is, data representing all
repeated judgments made by every available judge for each pair of
stimuli produced by each pair of individuals in the population, hence
its name. Conversely, Section~\ref{sec-theory-theoretical_SC} integrates
the assessment design features into what we call the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

SCM~\ref{fig-cj03_scm} illustrates the preliminary integration of
theoretical principles into the conceptual-population model of the
motivating example. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{hiabjk})\) and several related
variables. The subscripts \(h\) and \(i\) label the students who
authored the texts (individuals), while the indices \(a\) and \(b\)
represent the compared texts (stimuli). Moreover, the index \(j\) refers
to the judge, and \(k\) specifies the judgment number, accounting for
experimental conditions where judges perform the same judgment multiple
times, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}. Notably, the indexing system allows
comparisons between texts written by the same student (\(h = i\),
\(a \neq b\)) and between texts written by distinct students
(\(h \neq i\), where \(a = b\) is possible). However, it excludes cases
where judges compare a student's text to itself (\(h = i\), \(a = b\)),
as such comparison lacks practical meaning within the CJ framework.

In line with Thurstone's theory, the other variables present in
SCM~\ref{fig-cj03_scm} include the texts' discriminal processes
\((T_{ha}, T_{ib})\) and their discriminal difference \((D_{hiabjk})\)
(see Section~\ref{sec-thurstone_theory}). Additionally, based on the
arguments presented in Section~\ref{sec-theory-issue1b} and the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978},
the SCM~\ref{fig-cj03_scm} incorporates the judges' biases \((B_{kj})\).
Together with the outcome, these variables form the preliminary set of
endogenous variables,
\(V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}\). Finally,
the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\[
\begin{aligned}
  O^{cp}_{hiabjk} & := f_{O}(D_{hiabjk}) \\
  D_{hiabjk} & := f_{D}(T_{ha}, T_{ib}, B_{jk})
\end{aligned}
\]

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, showing the direction of causal
influence. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. Moreover, they often omit exogenous variables
for simplicity, following the \emph{standard representation}. However,
this section adopts the \emph{magnified representation} of DAGS, as
including exogenous variables can enhance clarity and reveal potential
issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

Figure~\ref{fig-cj03_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. This graph illustrates the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and the practical aspects of CJ experiments. It also
distinguishes between observable endogenous variables, such as the
outcome, and variables considered latent within the CJ framework, such
as the text's discriminal processes, their discriminal difference, and
the judges' biases.

Although specifying a system's data structure is not mandatory for using
SCMs and DAGs, defining one can sometimes enhance clarity and facilitate
the description of the system. Thus, to simplify the description of the
CJ system, we first define the vectors \(I\) and \(J\), along with the
matrices \(IA\) and \(JK\), as follows:

\begin{equation}\phantomsection\label{eq-mat12}{
I_{n_{I}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
IA_{(n_{I} \cdot n_{A}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
J_{n_{J}} = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
JK_{(n_{J} \cdot n_{K}) \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

where an element of \(I\) corresponds to a unique individual \(h\) or
\(i\), while each row of \(IA\) corresponds to a unique pairing of
individuals \(h, i\) with stimuli \(a, b\). Here, \(n_{I}\) represents
the number of individuals, while \(n_{A}\) denotes the number of stimuli
available per individual. Thus, vector \(I\) has a length of \(n_{I}\),
and matrix \(IA\) has a dimension of \((n_{I} \cdot n_{A})\) rows and
\(2\) columns. The vector and the matrix consist of scalars from the set
of natural numbers \(\mathbb{N}\).

Similarly, an element of \(J\) corresponds to a unique judge \(j\).
Moreover, each row of \(JK\) specifies a unique pairing of a judge \(j\)
with a judgment number \(k\). In this case, \(n_{J}\) indicates the
total number of judges, and \(n_{K}\) represents the number of judgments
each judge performs. As a result, vector \(J\) has a length of
\(n_{J}\), and matrix \(JK\) has a dimension of \((n_{J} \cdot n_{K})\)
rows and \(2\) columns. Like the previous structures, the vector and the
matrix consist of scalars from the set of natural numbers
\(\mathbb{N}\).

Additionally, we define the matrix \(RPUV\), where each row uniquely
pairs the corresponding rows from the matrices \(IA\) and \(JK\). This
matrix reflects the fundamental idea that the discriminal difference of
the stimuli becomes an observable outcome only through the judges'
perceptions. Thus, the matrix is defined as follows:

\begin{equation}\phantomsection\label{eq-mat3}{
RPUV_{{n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix} \in \mathbb{N}
}\end{equation}

This matrix also consists of scalars from the set of natural numbers
\(\mathbb{N}\) and has a dimension of
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\) rows and \(6\)
columns. The matrix's dimension results from the conceptual-population
model's assumption that researchers have access to data capturing all
repeated judgments made by every available judge for each pair of
stimuli produced by each pair of individuals in the population. Here,
the term \({n_{I} \cdot n_{A} \choose 2}\) represents the Binomial
coefficient, which quantifies the total number of individual-stimuli
comparison pairs within the conceptual population. The coefficient is
defined as follows:

\[
{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
\]

Readers can better visualize the structure of these vectors and matrices
by considering small values for the parameters corresponding to number
of individuals \((n_{I} = 5)\), the number of available stimuli per
individual \((n_{A} = 2)\), the total number of judges \((n_{J} = 3)\),
and the number of judgments each judge performs \((n_{K} = 3)\). Using
these values, the matrices described in equations (\ref{eq-mat12}) and
(\ref{eq-mat3}) take the following form:

\[
I_{5} = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
IA_{10 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
J_{3} = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
JK_{9 \times 2} = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
RPUV_{405  \times 6} = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

Thus, using the data structure defined by the matrices in equations
(\ref{eq-mat12}) and (\ref{eq-mat3}), we can now re-express
SCM~\ref{fig-cj03_scm} and DAG~\ref{fig-cj03_dag} in an equivalent
vectorized form, as shown in Figure~\ref{fig-cj04}. In this
representation, the outcome \(O^{cp}_{RPUV}\), the texts' discriminal
difference \(D_{RPUV}\), their discriminal processes \(T_{IA}\), and the
judges' biases \(B_{JK}\) are no longer scalar values but vectors. These
vectors capture all the values from the conceptual population.
Specifically, \(O^{cp}_{RPUV}\) and \(D_{RPUV}\) are vectors of length
\({n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\), while
\(T_{IA}\) and \(B_{JK}\) are vectors of length \((n_{I} \cdot n_{A})\)
and \((n_{J} \cdot n_{K})\), respectively.

@@

Building on the arguments developed in Section~\ref{sec-theory-issue2}
and the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model then includes two hierarchical
structural components. Each component defines how different observed or
latent variables influence the primary latent variable of interest
\citep{Everitt_et_al_2010}. This addition facilitates analyses that
integrate the trait measurement and hypothesis testing processes,
simultaneously accounting for the traits and their associated
uncertainties, thereby reducing bias and enhancing the measurement
precision.

Figure~\ref{fig-cj09} displays the two hierarchical structural
components in the motivating example. Specifically, the top branch of
DAG~\ref{fig-cj09_dag} depicts the first component. This component
describes how the students' writing-quality trait \((T_{I})\) and a set
of \emph{relevant} text-related variables \(X_{IA}\) causally influence
the texts' written-quality trait \(T_{IA}\). The DAG also highlights the
idiosyncratic errors \(e_{IA}\), which capture variations in the texts'
trait that neither the students' trait nor other relevant text-related
variables can explain. Here, \(T_{I}\) and \(e_{IA}\) are latent vectors
of length \(n_{I}\) and \((n_{I} \cdot n_{A})\), respectively.
Additionally, \(X_{IA}\) is an observed matrix with a dimension of
\((n_{I} \cdot n_{A})\) rows and \(p_{IA}\) columns.

Furthermore, the top branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} student-related variables \(X_{I}\) causally
influence the students' writing-quality trait \(T_{I}\). The DAG also
includes the idiosyncratic errors \(e_{I}\), which capture variations in
the the students' traits that the relevant student-related variables
cannot explain. Here, \(e_{I}\) is a latent vector of length \(n_{I}\),
and \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(p_{I}\)
columns.

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component. This component describes how the judges' bias
\(B_{J}\) and a set of \emph{relevant} judgment-related variables
\(Z_{JK}\) causally influence the biases associated with each text
\(B_{JK}\). The DAG also highlights the idiosyncratic errors \(e_{JK}\),
which capture variations in the texts' bias that neither the judges'
bias nor other relevant judgment-related variables can explain. In this
description, \(B_{J}\) and \(e_{JK}\) are latent vectors of length
\(n_{J}\) and \((n_{J} \cdot n_{K})\), respectively. Additionally,
\(Z_{JK}\) is an observed matrix with a dimension of
\((n_{J} \cdot n_{K})\) rows and \(p_{JK}\) columns.

Moreover, the bottom branch of DAG~\ref{fig-cj09_dag} describes how a
set of \emph{relevant} judge-related variables \(Z_{J}\) causally
influence the judges' bias \(B_{J}\). The DAG also accounts for
idiosyncratic errors \(e_{J}\), capturing the variations in the judges'
bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(p_{J}\) columns. Notably, all
variables and functions shown in SCM~\ref{fig-cj09_scm} and
DAG~\ref{fig-cj09_dag} constitute the set of endogenous variables \(V\)
and structural equations \(F\) for the conceptual-population model.

A cautious reader will notice that these descriptions rely on several
relevant variables. In this study, a \emph{relevant} or a
\emph{sufficient adjustment} set of variables (potentially an empty set)
refers to those that satisfy the \emph{backdoor criterion}
\citep[pp.~37]{Neal_2020}, that is, variables that block all non-causal
paths without introducing new ones
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}.

Incorporating these variables into the conceptual-population model
enables researchers to identify different estimands of interest and
their causal effects, even with observational data. This identification
works because conditioning on these variables ensures that the estimator
meets several critical properties, including confounding elimination
\citep{Morgan_et_al_2014}. Here, \emph{conditioning} means restricting
the statistical analysis to a subset of the population defined by the
values of a given variable \citep[pp.~32]{Neal_2020}.

Naturally, the validity of causal claims now hinges on the assumption
that these variables serve as a sufficient adjustment set, a property
that cannot be guaranteed for any variable. Nevertheless, as
\citet[pp.~150]{Kohler_et_al_2019} noted, drawing conclusions about the
real world from observed data inevitably requires assumptions, and these
requirements hold true for both observational and experimental data.

In summary,
